---
title: "Neosysiphus Niche Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Main functions}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

library(biomod2)
library(terra)

### Load dataset and variables

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}

# Load species occurrences (6 species available)
DataSpecies <- read.csv("Counting_Nesosisyphus.csv", sep = ";")

DataSpecies$loc <- paste(DataSpecies$long, DataSpecies$lat)

occur <- data.frame(matrix(ncol = length(unique(DataSpecies$Species))+2,
                           nrow = length(unique(DataSpecies$loc))))

colnames(occur) <- c("Lon", "Lat", unique(DataSpecies$Species))
for (i in 1:length(occur$Lon)){
  occur$Lon[i] <- DataSpecies$long[DataSpecies$loc == unique(DataSpecies$loc)[i]][1]
  occur$Lat[i] <- DataSpecies$lat[DataSpecies$loc == unique(DataSpecies$loc)[i]][1]
    
  occur[i,unique(DataSpecies$Species[DataSpecies$loc==unique(DataSpecies$loc)[i]])] <- 1
  
}

occur$`Nesosisyphus ` <- 1


# Select the name of the studied species
#Nesosisyphus rotundatus --> only 2 observation == too few
myRespName <- 'Nesosisyphus'

# Get corresponding presence/absence data
myResp <- as.numeric(occur[, myRespName])

# Get corresponding XY coordinates: X-longitude, Y-latitude
myRespXY <- occur[, c('Lon', 'Lat')]

# Load environmental variables extracted from BIOCLIM (bio1, bio5, bio6, bio12, bio16, bio17)

#BIO1 = Annual Mean Temperature
#BIO5 = Max Temperature of Warmest Month
#BIO6 = Min Temperature of Coldest Month
#BIO12 = Annual Precipitation
#BIO16 = Precipitation of Wettest Quarter
#BIO17 = Precipitation of Driest Quarter

fs <- list.files(path="./Mauritius_clim", pattern = "tif$", full.names = TRUE)
s <- raster::stack(fs)
clim <- raster::stackSave(s, "clim")

#myExpl <- rast(clim[[c(1, 15, 16, 4, 8, 9)]])
myExpl <- rast(clim[[c(1, 4)]])

#data("bioclim_current")
#myExpl <- rast(bioclim_current)
```


### Prepare data & parameters
summary(myBiomodData)

#### Format data (observations & explanatory variables)

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# # Format Data with true absences
myBiomodData <- BIOMOD_FormatingData(resp.var = myResp,
                                      expl.var = myExpl,
                                      resp.xy = myRespXY,
                                      resp.name = myRespName)
myBiomodData
plot(myBiomodData)

myBiomodData
summary(myBiomodData)
plot(myBiomodData)
```

#### Pseudo-absences extraction

Single or multiple set of pseudo-absences can be selected with the [`BIOMOD_FormatingData`](../reference/BIOMOD_FormatingData.html) function, which calls the [`bm_PseudoAbsences`](../reference/bm_PseudoAbsences.html) function to do so. More examples are presented on the [Secundary functions webpage](examples_2_secundaryFunctions.html).

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# # Transform true absences into potential pseudo-absences
myResp.PA <- ifelse(myResp == 1, 1, NA)
# Format Data with pseudo-absences : random method
myBiomodData.r <- BIOMOD_FormatingData(resp.var = myResp.PA,
                                       expl.var = myExpl,
                                       resp.xy = myRespXY,
                                       resp.name = myRespName,
                                       PA.nb.rep = 4,
                                       PA.nb.absences = 1000,
                                       PA.strategy = 'random')
myBiomodData.r
plot(myBiomodData.r)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Select multiple sets of pseudo-absences

# Transform true absences into potential pseudo-absences
myResp.PA <- ifelse(myResp == 1, 1, NA)

# Format Data with pseudo-absences : random method
myBiomodData.multi <- BIOMOD_FormatingData(resp.var = myResp.PA,
                                           expl.var = myExpl,
                                           resp.xy = myRespXY,
                                           resp.name = myRespName,
                                           PA.nb.rep = 4,
                                           PA.nb.absences = c(100, 100, 50, 50),
                                           PA.strategy = 'random')
myBiomodData.multi
summary(myBiomodData.multi)
plot(myBiomodData.multi)
```

#### Parameterize modeling options

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}


# Create default modeling options

default_options_binary <- bm_ModelingOptions(data.type = "binary", strategy = "default")
print(default_options_binary)


# # Part (or totality) of the print can be copied and customized
# # Below is an example to compute quadratic GLM and select best model with 'BIC' criterium
myBiomodOptions <- bm_ModelingOptions(
   GLM = list(type = 'quadratic',
              interaction.level = 0,
              myFormula = NULL,
              test = 'BIC',
              family = 'binomial',
              control = glm.control(epsilon = 1e-08,
                                    maxit = 1000,
                                   trace = FALSE)))
myBiomodOptions
# 
# # It is also possible to give a specific GLM formula
# myForm <- 'Sp277 ~ bio3 + log(bio10) + poly(bio16, 2) + bio19 + bio3:bio19'
# myBiomodOptions <- BIOMOD_ModelingOptions(GLM = list(myFormula = formula(myForm)))
# myBiomodOptions
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
### Model parameters can also be automatically tuned for your specific 
### dataset with an optimization algorithm. The tuning can however 
### be quite long. Duration for tuning all models sequentially 
### with default optimization settings :
### on 1 x 2.5 GHz processor: approx. 45 min tuning all models 
### on 8 x 2.5 GHz processor: approx. 15 min tuning all models 

 library(doParallel)
 cl <- makeCluster(8)
 doParallel::registerDoParallel(cl)
# 
 time.seq <- system.time(
   bm.tuning <- BIOMOD_Tuning(bm.format = myBiomodData, ME.env = myExpl, ME.n.bg = ncell(myExpl)))
# 
 stopCluster(cl)
# 
# plot(bm.tuning$tune.CTA.rpart)
# plot(bm.tuning$tune.CTA.rpart2)
# plot(bm.tuning$tune.RF)
# plot(bm.tuning$tune.ANN)
# plot(bm.tuning$tune.MARS)
# plot(bm.tuning$tune.FDA)
# plot(bm.tuning$tune.GBM)
# plot(bm.tuning$tune.GAM)
# 
# # Get tuned modeling options
myBiomodOptions <- bm.tuning$models.options
```

#### Cross-validation datasets

Several cross-validation methods are available and can be selected with the [`BIOMOD_Modeling`](../reference/BIOMOD_Modeling.html) function, which calls the [`bm_CrossValidation`](../reference/bm_CrossValidation.html) function to do so. More examples are presented on the [Secundary functions webpage](examples_2_secundaryFunctions.html).

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# # k-fold selection
# cv.k <- bm_CrossValidation(bm.format = myBiomodData,
#                            strategy = "kfold",
#                            nb.rep = 2,
#                            k = 3)
# 
# # stratified selection (geographic)
# cv.s <- bm_CrossValidation(bm.format = myBiomodData,
#                            strategy = "strat",
#                            k = 2,
#                            balance = "presences",
#                            strat = "x")
# head(cv.k)
# head(cv.s)
```


### Run modelisation

#### Single models

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Model single models
myBiomodModelOut <- BIOMOD_Modeling(bm.format = myBiomodData.multi,
                                    bm.options = default_options_binary,
                                    modeling.id = 'AllModels',
                                    CV.strategy = 'random',
                                    CV.nb.rep = 5,
                                    CV.perc = 0.8,
                                    var.import = 3,
                                    metric.eval = c('TSS','ROC'))
                                    # seed.val = 123)
                                    # nb.cpu = 8)
myBiomodModelOut

# Get evaluation scores & variables importance
get_evaluations(myBiomodModelOut)
get_variables_importance(myBiomodModelOut)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodModelOut)
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut, group.by = c('algo', 'algo'))
#bm_PlotEvalBoxplot(bm.out = myBiomodModelOut, group.by = c('algo', 'run'))
#bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut, group.by = c('expl.var', 'algo', 'algo'))
#bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut, group.by = c('expl.var', 'algo', 'run'))
#bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut, group.by = c('algo', 'expl.var', 'run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodModelOut, 
                      models.chosen = get_built_models(myBiomodModelOut)[c(1:3, 12:14)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodModelOut, 
                      models.chosen = get_built_models(myBiomodModelOut)[c(1:3, 12:14)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodModelOut, 
                      models.chosen = get_built_models(myBiomodModelOut)[3],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

#### Ensemble models

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Model ensemble models
myBiomodEM <- BIOMOD_EnsembleModeling(bm.mod = myBiomodModelOut,
                                      models.chosen = 'all',
                                      em.by = 'all',
                                      em.algo = c('EMmean', 'EMcv', 'EMci', 'EMmedian', 'EMca', 'EMwmean'),
                                      metric.select = c('TSS'),
                                      metric.select.thresh = c(0.7),
                                      metric.eval = c('TSS', 'ROC'),
                                      var.import = 3,
                                      EMci.alpha = 0.05,
                                      EMwmean.decay = 'proportional')
myBiomodEM

sessionInfo()


# Get evaluation scores & variables importance
get_evaluations(myBiomodEM)
get_variables_importance(myBiomodEM)

# Represent evaluation scores & variables importance
#bm_PlotEvalMean(bm.out = myBiomodEM, group.by = 'full.name')
#bm_PlotEvalBoxplot(bm.out = myBiomodEM, group.by = c('full.name', 'full.name'))
#bm_PlotVarImpBoxplot(bm.out = myBiomodEM, group.by = c('expl.var', 'full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM, group.by = c('expl.var', 'algo', 'merged.by.run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM, group.by = c('algo', 'expl.var', 'merged.by.run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodEM, 
                      models.chosen = get_built_models(myBiomodEM)[c(1, 6, 7)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodEM, 
                      models.chosen = get_built_models(myBiomodEM)[c(1, 6, 7)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodEM, 
                      models.chosen = get_built_models(myBiomodEM)[7],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```


#### Presence-only evaluation

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Evaluate models with Boyce index and MPA
myBiomodPO <- BIOMOD_PresenceOnly(bm.mod = myBiomodModelOut,
                                  bm.em = myBiomodEM)
myBiomodPO

# Evaluate models with Boyce index and MPA (using background data)
myBiomodPO <- BIOMOD_PresenceOnly(bm.mod = myBiomodModelOut,
                                  bm.em = myBiomodEM, 
                                  bg.env = values(myExpl))
myBiomodPO
```

### Project models

#### Single models

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Project single models
myBiomodProj <- BIOMOD_Projection(bm.mod = myBiomodModelOut,
                                  proj.name = 'Current',
                                  new.env = myExpl,
                                  models.chosen = 'all',
                                  metric.binary = 'all',
                                  metric.filter = 'all',
                                  build.clamping.mask = TRUE,
                                  compress = FALSE)
myBiomodProj
#plot(myBiomodProj)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Project SRE models
myBiomodProj_2 <- BIOMOD_Projection(bm.mod = myBiomodModelOut,
                                  proj.name = 'Current',
                                  new.env = myExpl,
                                  models.chosen =c('Nesosisyphus_PA1_RUN1_SRE','Nesosisyphus_PA1_RUN2_SRE', 'Nesosisyphus_PA3_RUN1_SRE', 'Nesosisyphus_PA3_RUN2_SRE', 'Nesosisyphus_PA1_RUN1_GLM','Nesosisyphus_PA1_RUN2_GLM'), 
                                  metric.binary = 'all',
                                  metric.filter = 'all',
                                  do.stack = FALSE)
myBiomodProj_2
plot(myBiomodProj_2)
```


#### Ensemble models

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Project ensemble models (from single projections)
myBiomodEMProj <- BIOMOD_EnsembleForecasting(bm.em = myBiomodEM, 
                                             bm.proj = myBiomodProj,
                                             models.chosen = 'all',
                                             metric.binary = 'all',
                                             metric.filter = 'all')
                                             
# Project ensemble models (building single projections)
myBiomodEMProj <- BIOMOD_EnsembleForecasting(bm.em = myBiomodEM,
                                             proj.name = 'CurrentEM',
                                             new.env = myExpl,
                                             models.chosen = 'all',
                                             metric.binary = 'all',
                                             metric.filter = 'all')
myBiomodEMProj
plot(myBiomodEMProj)
```